---
title: "Population Structure"
author: "SRK"
date: "2025-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Downloads")
```

library(ggplot2) # plotting
library(ggpubr) # plotting

## First, let's work on the genetic PCA:

COV <- as.matrix(read.table("RSBS_poly.cov")) # read in the genetic covariance matrix

PCA <- eigen(COV) # extract the principal components from the COV matrix

## How much variance is explained by the first few PCs?

var <- round(PCA$values/sum(PCA$values),3)

var[1:3]

# A "screeplot" of the eigenvalues of the PCA:

barplot(var, 
        xlab="Eigenvalues of the PCA", 
        ylab="Proportion of variance explained")

## Bring in the bam.list file and extract the sample info:

names <- read.table("RSBS_bam.list")
names <- unlist(strsplit(basename(as.character(names[,1])), split = ".sorted.rmdup.bam"))
split = strsplit(names, "_")
pops <- data.frame(names[1:113], do.call(rbind, split[1:113]))
names(pops) = c("Ind", "Pop", "Row", "Col")

## Plot the PCA using ggplot

data=as.data.frame(PCA$vectors)
data=data[,c(1:3)]
data= cbind(data, pops)

# A simple ggplot:

ggscatter(data, x = "V1", y = "V2",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE,
          main="Red spruce-Black spruce genetic PCA") +
  theme_bw(base_size = 13, base_family = "Times") +
  labs(x = paste0("PC1: (",var[1]*100,"%)"), y = paste0("PC2: (",var[2]*100,"%)"))



# A more customized plot

cols=c("#377eB8","#EE9B00","#0A9396","#94D2BD","#FFCB69","#005f73","#E26D5C","#AE2012", "#6d597a", "#7EA16B","#d4e09b", "gray70", "black", "darkgray", "purple")

ggscatter(data, x = "V1", y = "V2",
          color = "Pop",
          mean.point = TRUE,
          star.plot = TRUE) +
  theme_bw(base_size = 13, base_family = "Times") +
  theme(panel.background = element_blank(), 
        legend.background = element_blank(), 
        panel.grid = element_blank(), 
        plot.background = element_blank(), 
        legend.text=element_text(size=rel(.7)), 
        axis.text = element_text(size=13), 
        legend.position = "bottom") +
  labs(x = paste0("PC1: (",var[1]*100,"%)"), y = paste0("PC2: (",var[2]*100,"%)")) +
  scale_color_manual(values=c(cols), name="Source population") +
  guides(colour = guide_legend(nrow = 2))




## Next, we can look at the admixture clustering:

# import the ancestry scores (these are the .Q files)

q <- read.table("RSBS_poly.admix.2.Q", sep=" ", header=F)

K=dim(q)[2] #Find the level of K modeled

## order according to population code
ord<-order(pops[,2])

# make the plot:
barplot(t(q)[,ord],
        #col=cols[1:K],
        col=c("black","red"),
        space=0,border=NA,
        xlab="Populations",ylab="Admixture proportions",
        main=paste0("Red Spruce-Black Spruce Admixture: K=",K))
text(tapply(1:nrow(pops),pops[ord,2],mean),-0.05,unique(pops[ord,2]),xpd=T,cex=0.75,srt=45)
abline(v=cumsum(sapply(unique(pops[ord,2]),function(x){sum(pops[ord,2]==x)})),col=1,lwd=1.2)

